Biden Expected to Win 54% of Popular Vote in 2020 US Presidential Election

### 2020-11-03 outputs/paper/paper.Rmd
sad

output:
  bookdown::pdf_document2:
toc: FALSE
bibliography: references.bib
nocite: '@*'
link-citations: yes
---

```{r setup, include=FALSE}
library(haven)
library(tidyverse)
library(knitr)
library(palmerpenguins)
```
```{r loading, echo=FALSE}
# Load required data and models
cleaned_data <- data.frame(read.csv(file = "../../inputs/cleaned_acs.csv"))
```

@@

### Data collection
The post-stratification dataset was gathered from the American Community Survey (ACS), a project aiming to mitigate issues stemming from the Census’ 10-year intervals by providing an annualized version of data like that produced by the decennial census long form. The dataset used in this study is specifically that of the 2018 ACS data. The ACS data can be accessed at the IPUMS website. More details on attaining and cleaning data are found in 01-data_cleaning-post-strat.R in the scripts folder of the git repository. TODO

The target population, much like the Census, is essentially anyone who resides in a dwelling in the US. Following this, the sampling frame of the ACS is the Master Address File that is maintained by the US Census Bureau. (@cas2011) Created for the 2000 Census, it was originally based on the 1990 Address Control File and the United States Postal Service’s Delivery Sequence File. The maintaining and updating of this file is crucial to the efforts of the ACS and any other body that makes use of it. In addition, the ACS samples 2.5 percent of the population living in “Group Quarters”, which are non-housing units (eg. nursing homes, prisons, college dorms, etc.). In total, the 2018 ACS data contains about 3.2 million observations, sampled from across the country.

@@

As mentioned above, the variables we used were age (the age of the respondent), gender (what gender they identified as), race/ethnicity of the respondent, state (the state they lived in), and education (what their highest level of education was). We did not find any variables that were similar enough to the ones chosen. However, as the post-stratification data was a different dataset to that of the survey data, we had to transform the post-stratification data to match that, including general cleaning in the sense of converting non-numeric values to numeric, matching spelling/capitalization of certain responses, and constructing an age_groups variable from the ages given. For more information, please see 01-data_cleaning-post-strat.R in the scripts folder of the github repository.  TODO

@@

```{r, echo=FALSE}
coefficient <- matrix(c(1.40, 0.44, 0.53, 2.26, 1.39, 0.71, 0.07, 2.89, 0.75, 0.34, 0.08, 1.42, 2.38, 0.32, 1.75, 3.01, 
                        0.92, 0.32, 0.31, 1.54, 0.13, 0.29, -0.42, 0.70, -0.44, 0.06, -0.57, -0.32, -0.27, 0.21, -0.68, 0.13, -0.77, 0.20,-1.17,-0.38, -1.02,0.20,-1.42,-0.64,-0.94,0.21,-1.35,-0.55, -0.83,0.20,-1.25,-0.45, -0.74,0.22,-1.19,-0.32, 0.36,0.07,0.22,0.51), ncol=4, byrow=TRUE)
colnames(coefficient) <- c("Estimate", "Estimate Error", "l-95% CI", "u-95% CI")
rownames(coefficient) <- c("race_AsianChinese", "race_AsianJapanese", "race_AsianOther", "race_AfricanAmerican", "race_Someotherrace", "race_ethnicityWhite", "genderMale", "age_group29to38", "age_group39to48", "age_group49to58", "age_group59to68", "age_group69to78", "age_group79andabove", "education_PostSecondaryDegree")
coefficient <- as.table(coefficient)
kable(coefficient)
```
Table 6: Coefficients estimates and coefficient intervals given a confidence level of 95%

We consider the strength of our variables by looking at the coefficient estimates and how large the interval of the estimate is at a 95% confidence level (we calculate this by subtracting "u-95% CI" which stands for upper 95% confidence interval and "l-95% CI" which stands for lower 95% confidence interval). The output for the states are omitted in the interest of space (and because we noticed that the interval of their estimate is very large at a 95% confidence level is very large with absolute value of ~4 for each state, indicating that the estimate for each state is uncertain and thus state is not a strong predictor of voting). Although all variables listed in Table 6 have a smaller interval that state, a few variables stand out has having especially small confidence intervals. They are gender (interval of only 0.25) and education (interval of only 0.29). This indicates that the model is fairly certain in its estimate of the coefficients for gender and education. Therefore, these two variables likely have the strongest influence on the response variable because there is variability in the possible value of their coefficients at a confidence level of 0.95. Furthermore, the low error of these two variables reemphasizes our conclusion that age and education are the strongest determiners of 2020 voting. The other variables are stronger predictors than state but not as strong as age and education.

```{r, echo=FALSE}
cell_counts <- readRDS("../../inputs/cell_counts.RDS")
training_data <- readRDS("../../inputs/training_data.RDA")
model <- readRDS("../../model/4chains_3000iter_.rds")

@@

```{r raceproportions, fig.height=4, fig.cap="Proportion of forecasted votes for Biden by race", message=FALSE, echo=FALSE}
# following graphs are inspired by Monica Alexander 
# https://github.com/MJAlexander/marriage-name-change/blob/master/mncs_mrp.R
race_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(race_ethnicity), 
             color = "MRP estimate")) + 
  geom_point() +
  ylab("Proportion voting for Biden") + 
  xlab("Race") + 
  geom_point(data = training_data %>% 
               group_by(race_ethnicity, vote_biden) %>%
               dplyr::summarize(n = n()) %>% 
               group_by(race_ethnicity) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(race_ethnicity, prop, color = "ACS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", 
                                           "ACS raw data" = "red")) + 
  theme_bw(base_size = 14) + coord_flip()

@@

```{r stateproportions, fig.height=8,fig.width=8, fig.cap="Proportion of forecasted votes for Biden by state", message=FALSE, echo=FALSE}

@@

  ylab("Proportion voting for Biden") + 
  xlab("State") + 
  geom_point(data = training_data %>% 
               group_by(state, vote_biden) %>%
               dplyr::summarize(n = n()) %>% 
               group_by(state) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(state, prop, color = "ACS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "ACS raw data" = "red")) + 
  theme_bw() +
  coord_flip()
```

Similarly for age (Figure \@ref(fig:ageproportions)), there showed a general trend that the older population are less likely to vote for Trump compared to Biden, with 18-28 years old population projected to vote for Biden almost 70% of the time and just over 45% of the 19 to 58 age group predicted to vote for Biden.

```{r ageproportions, fig.height=4,fig.width=8, fig.cap="Proportion of forecasted votes for Biden by age group.", message=FALSE, echo=FALSE}

age_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(age_group), color = "MRP estimate")) + 
  geom_point() +
  ylab("Proportion voting for Joe Biden") + 
  xlab("Age") + 
  geom_point(data = training_data %>% 
               group_by(age_group, vote_biden) %>%
               dplyr::summarize(n = n()) %>% 
               group_by(age_group) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(age_group, prop, color = "ACS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "ACS raw data" = "red")) + 
  theme_bw(base_size = 14) +
  coord_flip()

@@

Finally in Figures \@ref(fig:genderprop) and \@ref(fig:ageproportions) below, we show that Females are more likely to vote for Biden compared to Males, and a larger portion of individuals with at least some college education are projected to vote for Biden in comparison to High School educated and below.

```{r genderprop, fig.height=4,fig.width=8, fig.cap="Proportion of forecasted votes for Biden by gender.", message=FALSE, echo=FALSE}
gender_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(gender), color = "MRP estimate")) + 
  geom_point() +

@@

```{r educationprop, fig.height=4,fig.width=8, fig.cap="Proportion of forecasted votes for Biden by education.", message=FALSE, echo=FALSE}
# education_res$age_group <- as.factor(education_res$age_group)

@@

For the US 2020 election between Biden and Trump, we forecast Biden to win the popular vote by 54% to 46% of the votes \@ref(fig:resultspop).

```{r resultspop, fig.height=4,fig.width=8, fig.cap="Forecasted popular vote results.", message=FALSE, echo=FALSE}

@@

data.frame(Candidates, Percentage) %>% 
  ggplot(aes(x = Candidates, 
             y= Percentage)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage*100), "%")),
            color="white",
            position = position_stack(vjust = 0.5))
```

# Discussion
Figure \@ref(fig:ageprop) features the trend of the younger voters (between the ages of 18 and 38) to vote left-wing. There are many different reasons this can be the case. Younger people generally tend to be left wing (pew 1). In addition, given the current state of the economy, combined with the fact that in general millennials and Gen-Xers are poorer than their parents, many younger voters are in disillusion with the current government, leading to vote against the Republican Party (mckinsey).

@@

@@

While our model using multilevel regression with post-stratification was able to provide a election result that resembles many other polling predictions (Biden winning popular vote slightly ahead of Trump), our model had many weaknesses that we would like to address and improve in the future. First and foremost, we only predicted the the votes towards either Biden or Trump, and did not consider the possibilty that people may vote for other candidates or people may not vote in this election. While justified as the vast majority of votes entered will go towards Biden and Trump, it will be insightful to analyze which demographics are likely to not vote for the two main parties, and which demographics are likely to not vote at all (maybe younger people are more apathetics to voting, or certain minorities always vote). There were also many articles we found indicating variables such as religion and family status could play a key indication of how individuals vote, but due to restrictions of the datasets, we did not have the information available. We also originally planned on creating a time series analysis on the Nationscape dataset in order to gain some insight in the data we used to train our model which we would like to include in the future. Furthermore, we would like to further explore various variables and factors of our model, in order to get a more significant accuracy.

# Appendix {-}

>>>>

output:
  bookdown::pdf_document2:
toc: FALSE
bibliography: references.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(haven)
library(tidyverse)
library(knitr)
```
```{r loading, echo=FALSE}
# Load required data and models
cleaned_data <- data.frame(read.csv(file = "../../inputs/cleaned_acs.csv"))
```

@@

### Data collection
The post-stratification dataset was gathered from the American Community Survey (ACS), a project aiming to mitigate issues stemming from the Census’ 10-year intervals by providing an annualized version of data like that produced by the decennial census long form. The dataset used in this study is specifically that of the 2018 ACS data. The ACS data can be accessed at the IPUMS website. More details on attaining and cleaning data are found in 01-data_cleaning-post-strat.R in the scripts folder of the git repository.

The target population, much like the Census, is essentially anyone who resides in a dwelling in the US. Following this, the sampling frame of the ACS is the Master Address File that is maintained by the US Census Bureau. Created for the 2000 Census, it was originally based on the 1990 Address Control File and the United States Postal Service’s Delivery Sequence File. The maintaining and updating of this file is crucial to the efforts of the ACS and any other body that makes use of it. In addition, the ACS samples 2.5 percent of the population living in “Group Quarters”, which are non-housing units (eg. nursing homes, prisons, college dorms, etc.). In total, the 2018 ACS data contains about 3.2 million observations, sampled from across the country.

@@

As mentioned above, the variables we used were age (the age of the respondent), gender (what gender they identified as), race/ethnicity of the respondent, state (the state they lived in), and education (what their highest level of education was). We did not find any variables that were similar enough to the ones chosen. However, as the post-stratification data was a different dataset to that of the survey data, we had to transform the post-stratification data to match that, including general cleaning in the sense of converting non-numeric values to numeric, matching spelling/capitalization of certain responses, and constructing an age_groups variable from the ages given. One more thing of importance is that we chose not to include data for those aged 88 and above, as their data interfered with the model and their contribution overall is not very significant. For more information, please see 01-data_cleaning-post-strat.R in the scripts folder of the github repository.

@@

```{r, echo=FALSE}
summary(model_education)
```

```{r, echo=FALSE}
#load pre preprocessed items, created through the scripts
cell_counts <- readRDS("../../inputs/cell_counts.RDS")
training_data <- readRDS("../../inputs/training_data.RDA")
model <- readRDS("../../model/4chains_3000iter_.rds")

@@

```{r raceproportions, fig.height=4,fig.width=8, fig.cap="Proportion of Biden Votes by race", message=FALSE, echo=FALSE}

# following graphs are inspired by Monica Alexander 
# https://github.com/MJAlexander/marriage-name-change/blob/master/mncs_mrp.R

race_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(race_ethnicity), 
             color = "MRP estimate")) + 
  geom_point() +
  ylab("Proportion voting for Biden") + 
  xlab("Race") + 
  geom_point(data = training_data %>% 
               group_by(race_ethnicity, vote_biden) %>%
               dplyr::summarize(n = n()) %>% 
               group_by(race_ethnicity) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(race_ethnicity, prop, color = "ACS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", 
                                           "ACS raw data" = "red")) + 
  theme_bw(base_size = 14) + coord_flip()

@@

```{r stateproportions, fig.height=8,fig.width=8, fig.cap="Proportion of Biden Votes by state", message=FALSE, echo=FALSE}

@@

  ylab("Proportion voting for Biden") + 
  xlab("State") + 
  geom_point(data = training_data %>% 
               group_by(state, vote_biden) %>%
               dplyr::summarize(n = n()) %>% 
               group_by(state) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(state, prop, color = "ACS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "ACS raw data" = "red")) + 
  theme_bw() +
  coord_flip()

@@

Similarly for age, there showed a general trend that the older population are less likely to vote for Trump compared to Biden, with 18-28 years old population projected to vote for Biden almost 70% of the time and just over 45% of the 19 to 58 age group predicted to vote for Biden (Figure \@ref(fig:ageproportions))
```{r ageproportions, fig.height=4,fig.width=8, fig.cap="Proportion of Biden votes within age group.", message=FALSE, echo=FALSE}

age_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(age_group), color = "MRP estimate")) + 
  geom_point() +
  ylab("Proportion voting for Joe Biden") + 
  xlab("Age") + 
  geom_point(data = training_data %>% 
               group_by(age_group, vote_biden) %>%
               dplyr::summarize(n = n()) %>% 
               group_by(age_group) %>% 
               mutate(prop = n/sum(n)) %>% 
               filter(vote_biden==1), 
             aes(age_group, prop, color = "ACS raw data")) +
  scale_color_manual(name = "", values = c("MRP estimate" = "black", "ACS raw data" = "red")) + 
  theme_bw(base_size = 14) +
  coord_flip()
```

Finally in Figures \@ref(fig:genderproportions) and \@ref(fig:educationproportions) below, we show that Females are more likely to vote for Biden compared to Males, and a larger portion of individuals with at least some college education are projected to vote for Biden in comparison to High School educated and below.

```{r genderproportions, fig.height=4,fig.width=8, fig.cap="Proportion of Biden votes within age group.", message=FALSE, echo=FALSE}
gender_res %>% 
  ggplot(aes(y = mean, x = forcats::fct_inorder(gender), color = "MRP estimate")) + 
  geom_point() +

@@

```{r educationproportions, fig.height=4,fig.width=8, fig.cap="Proportion of Biden votes within age group.", message=FALSE, echo=FALSE}
# education_res$age_group <- as.factor(education_res$age_group)

@@

For the US 2020 election between Biden and Trump, we forecast Biden to win the popular vote by 54% to 46% of the votes.

```{r}

@@

data.frame(Candidates, Percentage) %>% 
  ggplot(aes(x = Candidates, 
             y= Percentage)) +
  geom_bar(stat = "identity")
```

# Discussion
Figure 1 features the trend of the younger voters (between the ages of 18 and 38) to vote left-wing. There are many different reasons this can be the case. Younger people generally tend to be left wing (pew 1). In addition, given the current state of the economy, combined with the fact that in general millennials and Gen-Xers are poorer than their parents, many younger voters are in disillusion with the current government, leading to vote against the Republican Party (mckinsey).

@@

## States

Overall, as shown in results, the states are mostly split between Trump and Biden. Certain states such as Vermont and Arizona heavily favour Biden and Trump respectively. Many of the states that are currently split will be a major focus for the presidential candidates as gaining these votes can be crucial to winning the election. As such these states will likely have received a lot of attention in the past few weeks.

If we had the ability to allocate a greater budget to polling companies, we assuredly could achieve a better estimate. The key improvement would be to ask more people and ask more questions, such as religion. With a greater number of questions we could subsequently create more cells to work with, and with increasing the range of respondents stratification can occur.

@@

While our model using multilevel regression with post-stratification was able to provide a election result that resembles many other polling predictions (Biden winning popular vote slightly ahead of Trump), our model had many weaknesses that we would like to address and improve in the future. First and foremost, we only predicted the the votes towards either Biden or Trump, and did not consider the possibilty that people may vote for other candidates or people may not vote in this election. While justified as the vast majority of votes entered will go towards Biden and Trump, it will be insightful to analyze which demographics are likely to not vote for the two main parties, and which demographics are likely to not vote at all (maybe younger people are more apathetic to voting, or certain minorities always vote). There were also many articles we found indicating variables such as religion and family status could play a key indication of how individuals vote, but due to restrictions of the datasets, we did not have the information available. We also originally planned on creating a time series analysis on the Nationscape dataset in order to gain some insight in the data we used to train our model which we would like to include in the future. Furthermore, we would like to further explore various variables and factors of our model, in order to get a more significant accuracy.
